6.867 Problem Set 1

### 2015-10-03 HW1/tex/HW1.tex
[homework1] reorganize stuff.

\begin{document}
	\title{6.867 Problem Set 1}
	\date{October 1, 2015}
	\maketitle

\subsubsection*{Problem 1}

We implemented a rather naive gradient descent in Python using an algorithm similar to the pseudocode below. The function is written in such a way that it can accept arbitrary scalar functions $f$ and their gradients, $\nabla f(x)$, where $x$ can be a vector of arbitrary length, $n$. The goal of this function is to try and return $\arg\min_x {f(x)}$.

@@

In order to understand the effect of various hyperparameters (our initial guess, the step size, and the convergence criteria) of the algorithm, we compared the results of our gradient descent routine given a number of different parameter combinations for two different functions: $f(x) = \sum_i^n x_i^2$ and $f(x) = \sum_i^n \frac{1}{100}x_i^2 - \cos(x)$. The first of these is a convex function, whereas the second is quite non-convex, with many local minima. The results of our gradient descent algorithm for numerous values are found below in Table 1.

@@

The non-convex function, on the other hand, is much more sensitive to the hyperparameters we specify. The most notable departure from the convex function is the influence of our initial guess. It is much easier for our gradient descent algorithm to get trapped in local minima, and this is clearly illustrated in our results. While this is the most obvious difference, note that changes in both the learning rate and the convergence criteria also lead to non-trivially different guesses for $\arg\min_x {f(x)}$.

Generally speaking, choice of starting value is only in when were are dealing with non-convex functions as different starting values may lead to different local extremums. Though if we randomly initialized good starting values, it decreases the number of iterations necessary for convergence. Tuning the step size alters the performance of the algorithm, while tuning the convergence threshold alters the accuracy of the algorithm.

To elaborate on this a little, for any given convergence threshold, increasing step size will decrease the number of iterations required for convergence. However, we need make sure that the step isn't so large that the algorithm gets caught in a loop where its constantly overshooting the minimum. On the other hand, for a given step size, decreasing the convergence threshold increases accuracy of the algorithm, but at the cost of performance as it'll take additional iterations to reach the higher level of precision. Here, we want to make sure that the convergence threshold isn't so small that we can never converge.

Unfortunately, it might not always be convenient to specify the gradient of our scalar function, $f(x)$, explicitly. This might be because it is complicated to compute, or because no closed form solution exists. In order to get around this issue, we can approximate the gradient of the function using central differences. We approximate the $i$th element of the gradient, $\nabla_i$, as

@@

\noindent where $h$ is some small constant (we use $h = .001$).

We see that the performance in the calculation of the gradient is identical between the analytical approach and the central difference approach. That's great! Now let's see how our gradient descent algorithm compares to two more sophisticated optimizers - BFGS and Conjugate Gradient (CG). We compare both the number of iterations and function evaluations. We'll consider our algorithm in two forms - one where we pass in an explicit gradient function, and another where we estimate the gradient using the central difference method. All benchmarking is done with the same starting point: [5, -3, 7, 8]. $f_1(x) =  \sum_i^n x_i^2$ and $f_2(x) = \sum_i^n x_i^{\frac{2}{100}} + \cos(x)$. We see that when passed an analytical solution for $\nabla f(x)$, our algorithm requires about as many function calls, although many more iterations than either BFGS or CG. If we use the central difference approximation for $\nabla f(x)$ instead, our algorithm requires many more iterations and function calls.

\begin{table}
\captionof{table}{Performance of gradient descent given various algorithm hyperparameters} 
\begin{tabular}{llllll}
\toprule
{} &                        $Guess_{i}$ &      step &   Thresh. &    $Guess_{f}$ ($f(x) = \sum_i^n x_i^2$) & $Guess_{f}$ ($f(x) = \sum_i^n \frac{1}{100} x_i^2 - \cos(x)$) \\
\midrule
&    [5.0, -3.0, 7.0, 8.0] &    0.3 &    0.1 &   [0.020, -0.012, 0.029, 0.033] &    [6.150, -0.314, 6.165, 6.176] \\
&    [5.0, -3.0, 7.0, 8.0] &    0.3 &  0.001 &   [0.001, -0.001, 0.002, 0.002] &    [6.159, -0.025, 6.160, 6.161] \\
&    [5.0, -3.0, 7.0, 8.0] &    0.3 &  1e-05 &   [0.000, -0.000, 0.000, 0.000] &    [6.160, -0.003, 6.160, 6.160] \\
&    [5.0, -3.0, 7.0, 8.0] &  0.003 &    0.1 &   [1.172, -0.703, 1.641, 1.876] &    [5.005, -2.999, 6.995, 7.993] \\
&    [5.0, -3.0, 7.0, 8.0] &  0.003 &  0.001 &   [0.118, -0.071, 0.165, 0.188] &    [6.092, -0.932, 6.200, 6.267] \\
&    [5.0, -3.0, 7.0, 8.0] &  0.003 &  1e-05 &   [0.012, -0.007, 0.017, 0.019] &    [6.092, -0.932, 6.200, 6.267] \\
 &  [-10.0, 4.0, 1.0, -5.0] &    0.3 &    0.1 &  [-0.041, 0.016, 0.004, -0.020] &  [-12.059, 6.070, 0.014, -6.139] \\
&  [-10.0, 4.0, 1.0, -5.0] &    0.3 &  0.001 &  [-0.003, 0.001, 0.000, -0.001] &  [-12.294, 6.152, 0.001, -6.158] \\
 &  [-10.0, 4.0, 1.0, -5.0] &    0.3 &  1e-05 &  [-0.000, 0.000, 0.000, -0.000] &  [-12.315, 6.159, 0.000, -6.159] \\
 &  [-10.0, 4.0, 1.0, -5.0] &  0.003 &    0.1 &  [-2.388, 0.955, 0.239, -1.194] &  [-10.002, 4.004, 0.995, -5.005] \\
 &  [-10.0, 4.0, 1.0, -5.0] &  0.003 &  0.001 &  [-0.240, 0.096, 0.024, -0.120] &  [-11.748, 5.885, 0.063, -6.076] \\
 &  [-10.0, 4.0, 1.0, -5.0] &  0.003 &  1e-05 &  [-0.024, 0.010, 0.002, -0.012] &  [-11.848, 5.938, 0.051, -6.092] \\
 &   [1.0, -1.0, -3.0, 2.0] &    0.3 &    0.1 &  [0.010, -0.010, -0.031, 0.020] &   [0.007, -0.007, -0.314, 0.024] \\
 &   [1.0, -1.0, -3.0, 2.0] &    0.3 &  0.001 &  [0.001, -0.001, -0.002, 0.001] &   [0.001, -0.001, -0.025, 0.002] \\
 &   [1.0, -1.0, -3.0, 2.0] &    0.3 &  1e-05 &  [0.000, -0.000, -0.000, 0.000] &   [0.000, -0.000, -0.003, 0.000] \\
 &   [1.0, -1.0, -3.0, 2.0] &  0.003 &    0.1 &  [0.736, -0.736, -2.207, 1.471] &   [0.995, -0.995, -2.999, 1.994] \\
 &   [1.0, -1.0, -3.0, 2.0] &  0.003 &  0.001 &  [0.073, -0.073, -0.220, 0.147] &   [0.051, -0.051, -0.932, 0.143] \\
 &   [1.0, -1.0, -3.0, 2.0] &  0.003 &  1e-05 &  [0.007, -0.007, -0.022, 0.015] &   [0.051, -0.051, -0.932, 0.143] \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\captionof{table}{Performance in gradient calculation for analytical and central difference approaches} 
\begin{tabular}{llllll}
\toprule
{} &   $x_{i}$  &  function &   $\nabla_{analytical}$  & $\nabla_{approximate}$ \\
\midrule
&    [5.0, -3.0, 7.0, 8.0] & $f(x) = \sum_i^n x_i^2$ &  [ 10.  -6.  14.  16.] & [ 10.  -6.  14.  16.]\\
&    [5.0, -3.0, 7.0, 8.0] & $f(x) = \sum_i^n \frac{1}{100} x_i^2 - \cos(x)$ & [-0.8589 -0.2011  0.797   1.1494] &[-0.8589 -0.2011  0.797   1.1494] \\
&    [-10.   4.   1.  -5.] & $f(x) = \sum_i^n x_i^2$ &   [-20.   8.   2. -10.] &  [-20.   8.   2. -10.]\\
&    [-10.   4.   1.  -5.] & $f(x) = \sum_i^n \frac{1}{100} x_i^2 - \cos(x)$ &[ 0.344  -0.6768  0.8615  0.8589]&[ 0.344  -0.6768  0.8615  0.8589] \\
&    [ 1. -1. -3.  2.] & $f(x) = \sum_i^n x_i^2$ &  [ 10.  -6.  14.  16.] & [ 10.  -6.  14.  16.]\\
&    [ 1. -1. -3.  2.] & $f(x) = \sum_i^n \frac{1}{100} x_i^2 - \cos(x)$ & [ 0.8615 -0.8615 -0.2011  0.9493] & [ 0.8615 -0.8615 -0.2011  0.9493] \\
\bottomrule
\end{tabular}
\end{table}

@@

\subsubsection*{Problem 2}

With problem 2, we first derive the maximum-likelihood estimator for the weight vector $\theta$. Note that if we assume Gaussian errors, that is $Y \sim N(X\theta, \sigma^2)$, then our maximum-likelihood estimator is equivalent to the OLS estimator. Hence:
\begin{equation*}
	\hat{\theta}_{ML}=\hat{\theta}_{OLS}=(\Phi^T \Phi)^{-1}\Phi^T Y
\end{equation*}
where $\Phi$ denotes the polynomial basis expansion of $X$.

@@

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=2in]{Meq0.png}
		\caption*{$M=0$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=2in]{Meq1.png}
		\caption*{$M=1$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=2in]{Meq3.png}
		\caption*{$M=3$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=2in]{Meq9.png}
		\caption*{$M=9$}
	\end{minipage}
	\caption{Plots of polynomials having various orders $M$, in red, fitted to the the observed points. The green curve shows the DGF $\sin(2\pi x)$.}
\end{figure}

@@

The sum of squared errors of a linear polynomial basis model is defined by: 
\begin{equation*}
	SSE=(\Phi \theta-Y)^T(\Phi \theta-Y) 
\end{equation*}
which has gradient:
\begin{equation*}
	\nabla SSE = 2\Phi^T(\Phi \theta-Y)
\end{equation*}
We tested the accuracy to this closed form solution using our central differences approximator. The analytical and numerical gradients were essentially identical across many different points, even at order $M=9$. This is exactly what we want!

We then applied gradient descent to the SSE function using the analytical gradient to see if we could replicate the plots in Bishop. Optimizing our hyperparameters got increasingly difficult as we increased the order $M$. For $M=O$, we merely needed to tune the step size and we could achieve consistent convergence to the Bishop plot across many different starting values. At $M=1$, we needed to change both the step size and increase the number of maximum iterations before we consistently converged to the optimum. At $M=3$, we also had to tune the convergence threshold to be several order of magnitudes smaller than it was before, and further increase the number of max interations to get consistent convergence. Finally $M=9$, regardless of how we tried to tune the hyperparameters, we couldn't manage to find a set of hyperparameters that consistently converged, in fact, even scipy's BFGS optimizer failed to consistently produce a plot similar to Bishop's.

@@

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=2in]{Meq0T.png}
		\caption*{$M=0$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=2in]{Meq1T.png}
		\caption*{$M=1$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=2in]{Meq3T.png}
		\caption*{$M=3$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=2in]{Meq9T.png}
		\caption*{$M=9$}
	\end{minipage}
	\caption{Plots of trigonometric basis expansions with various orders $M$, in red, fitted to the the observed points. The green curve shows the DGF $\sin(2\pi x)$.}
\end{figure}

@@

\subsubsection*{Problem 3}

To begin problem 3, we implemented the closed form solution of Ridge Regression in Python, making use of Scipy and Numpy. Note that if you want to test cases where the number of features is greater than the number of data points (something that a well-trained teacher of machine should be skeptical to do), it is useful to use the pseudo-inverse, rather than the inverse to fit your parameters. We learned this the long way after debugging code for hours.

@@

Given the data in Bishop 1.4 (which can be seen below), we attempt to fit polynomials of various orders, $M$.

\begin{center}
\includegraphics[scale=.3]{Bishop_14_Data.png}
\end{center}

The plot below shows the mean squared error (MSE) that our ridge regression fit gives for values of $M$ between 1 and 5, for values of $\lambda$ ranging from $10^0 -1$ to $10^2$. Note that in the case where $\lambda = 0$, we have $OLS$ regression, and as $M$ increases, the model is free to overfit our data. While the MSE here is low, if we evaluated this on a new data point, our model would likely perform very poorly. As $\lambda$ gets higher, we penalize the model for adding additional large weights to $\hat{\theta}$, but our $MSE$ stays relatively large. However, these "large" MSE models are likely to perform better on a new data point!

\begin{center}
\includegraphics[scale=.4]{MSE_Lambda_Bishop.png}
\end{center}

In order to avoid overfitting, but in order to make sure that we have chosen values of $\lambda$ and $M$ (and in general, hyperparameters) that make sense, it is common to split the data into a training set and a test set. We will use the closed-form solution for ridge regression to calculate $\hat{\theta}$ on our training set. We will then use that $\hat{\theta}$ to attempt to predict values of $\mathbf{y}$ for the test data. By choosing the $\hat{\theta}$ that minimizes MSE on the test data, but is trained on the training data, we are ensuring that we have not overfit our model to the data it is trained on. As one final step in the model selection process, we can check the performance of our final $\hat{\theta}$ on a validation set, to make sure that performance is still good (and we have not overfit for performance on our test set).

@@

Let's now use this methodology on a real dataset, as opposed to a toy dataset. The dataset that we will use was compiled by Kirsztian Buza at Budapest University of Technology and Economics. It includes 280 numerical values describing a number of blog posts, such as day of week, number of parent pages, and number of comments received within 24 hours. We are hoping to predict the number of comments it will receive in the next 24 hours. We are given this data split into training, validation, and test datasets, so we will once again perform a grid search to find the optimal value of $\lambda$. In this particular case, we are not regressing a polynomial of arbitrary order, so we do not need to test different values of $M$.

\begin{center}
\includegraphics[scale=.4]{MSE_comparison.png}
\end{center}

We find that for very large $\lambda$ (as large as $10^6$, $MSE$ continues to decrease, although it is always close to 900. The MSE for a uniform predictor that predicts the average of $y$, $\hat{y}$, has an MSE of 1453, so the performance of our model is still better. However, we see that for values of $\lambda$ higher than $10^6$ or so, the value of the MSE increases, and eventually approaches (and eclipses) the MSE of our uniform estimator.

This isn't surprising - the blog feedback dataset is large, with 10,480 observations in the test set. If we interpret the value of $\lambda$ as strength of our prior, it makes sense that it does not have a substantial effect until it is significantly larger than the size of our data. Furthermore, the inability of our model to get very low MSE for low $\lambda$ speaks both to the fact that the data has a much higher dimension than our training data, and also suggests that the simple linear model we are using here may not be a very good fit for the data.

\begin{center}
\includegraphics[scale=.4]{BlogFeedbackRegressionMSE.png}
\end{center}

\subsubsection*{Problem 4}
For problem 4, we are working with a highly sparse dataset so we can compare the differences between LASSO and ridge regression. We first start by determining the optimal penalty $\lambda$ that minimizes the MSE of test set. Our optimal $\lambda$, 0.1, results in an MSE of $0.4395$. When we disable the $L_2$ regularizer, our ridge regression produces the same estimated parameters as OLS which results in a MSE of $4.0979$. Separately, we determine that $\lambda = .9$ is optimal LASSO penalty. This resulted in a MSE of $0.4524$. We plotted our 3 obtained fits - OLS, Ridge with $\lambda=0.1$, and LASSO with $\lambda=0.9$ - against the the true DGF:

\begin{center}
\includegraphics[width=4.5in]{plot4}	
\end{center}

While its seems the ridge regression seems to fit the true DGF slightly better than LASSO, both these methods vastly outperform standard OLS. We further plotted the estimated optimal weight vectors from our 3 models against the true $w$:

\begin{center}
\includegraphics[width=4.5in]{Weight_Distributions.png}	
\end{center}

We see that the L1 and L2 regularizers are working as intended! Our ridge regression estimates are considerably smaller than our OLS estimates. Furthermore, our we see sparse estimation from LASSO. This sparsity results from the differences between the L1 and L2 penalty. In ridge regression, we are want to minimize the squared error, subject to the constraint that the sum of the weights squared is less than some constant. Geometrically, we can think of this constraint as a sphere of some distance around the origin. In LASSO, we are minimizing the squared error subject to the constraint the the sum of the absolute values of the weights is less than some constant. The L1 constraint can be thought of as a diamond (or regular octahedral) around the origin.

Now consider sets weights that lead to the the same squared error. These sets which can represented as ellipses with the OLS estimates at their center, as that's the point the minimizes squared error. In Ridge regression, we are looking for the point where one of these ellipses touches the sphere around the origin, while in LASSO we're looking for the ellipse that touched the diamond. Since the diamond has shaper points, its fairly likely that the ellipse intersects the diamond at one of its points, we represents some weight begin equal to 0.

>>>>

\begin{document}
	\title{6.867 Problem Set 1}
	\maketitle

\subsubsection*{Implementing Gradient Descent}

We implemented a rather naive gradient descent algorithm in Python using an approach similar to the pseudocode below. The function is written in such a way that it can accept arbitrary scalar functions $f$ and their gradients, $\nabla f(x)$, where $x$ can be a vector of arbitrary length, $n$. If $\nabla f(x)$ is not specified, we approximate it using the central differences. The goal of this function is to try and return $\arg\min_x {f(x)}$.

@@

In order to understand the effect of various hyperparameters (our initial guess, the step size, and the convergence criteria) of the algorithm, we compared the results of our gradient descent routine given a number of different parameter combinations for two different functions: $f(x) = \sum_i^n x_i^2$ and $f(x) = \sum_i^n \frac{1}{100}x_i^2 - \cos(x)$. The first of these is a convex function, whereas the second is quite non-convex, with many local minima. A sample of the results of our gradient descent algorithm for numerous values are found below in Table 1.

@@

The non-convex function, on the other hand, is much more sensitive to the hyperparameters we specify. The most notable departure from the convex function is the influence of our initial guess. It is much easier for our gradient descent algorithm to get trapped in local minima, and this is clearly illustrated in our results. While this is the most obvious difference, note that changes in both the learning rate and the convergence criteria also lead to non-trivially different guesses for $\arg\min_x {f(x)}$. Note that we do put a cap on the number of iterations allowable, so it is also possible that our non-convex function simply fails to converge, although this doesn't appear to happen in any of the cases illustrated in this text.

Generally speaking, choice of starting value is only important when dealing with non-convex functions, as different starting values may lead to different local extremums. That being said, even for convex functions, good starting values decrease the number of iterations necessary for convergence. Tuning the step size alters the performance of the algorithm, while tuning the convergence threshold alters the accuracy of the algorithm.

It's worth elaborating on the nuances of picking a step size and convergence threshold. For any given convergence threshold, increasing step size will decrease the number of iterations required for convergence. However, we need make sure that the step isn't so large that the algorithm gets caught in a loop where its constantly overshooting the minimum. On the other hand, for a given step size, decreasing the convergence threshold increases accuracy of the algorithm, but it will take additional iterations to reach this higher level of precision. We want to make sure that the convergence threshold isn't so small that we can never converge.

Unfortunately, it might not always be convenient to specify the gradient of our scalar function, $f(x)$, explicitly. It may be complicated to compute, or there may not be a closed form solution. In order to get around this issue, we can approximate the gradient of the function using central differences. We approximate the $i$th element of the gradient, $\nabla_i$, as

@@

\noindent where $h$ is some small constant (we typically use $h = .001$).

The central difference approximation of the gradient is equivalent to the analytical solution for both the convex and non-convex function. That's great! We also want to know how our gradient descent algorithm compares to more sophisticated optimizers - BFGS and Conjugate Gradient (CG). We compare both the number of iterations and function evaluations required to calculate the minimum of the function. We'll check both the case where we provide an expression for the gradient, and the case where we estimate the gradient. All benchmarking is done with the same starting point: [5, -3, 7, 8]. $f_1(x) =  \sum_i^n x_i^2$ and $f_2(x) = \sum_i^n x_i^{\frac{2}{100}} + \cos(x)$. We see that when passed an analytical solution for $\nabla f(x)$, our algorithm requires about as many function calls, although many more iterations than either BFGS or CG. If we use the central difference approximation for $\nabla f(x)$ instead, our algorithm requires many more iterations and function calls. Later on in this paper, we'll make use of this algorithm to solve some machine learning problems.

\begin{table}
\captionof{table}{Performance of gradient descent given various algorithm hyperparameters} 
\begin{tabular}{llllll}
\toprule
{} &                        $Guess_{i}$ &      step &   Thresh. &    $Guess_{f}$ ($f(x) = \sum_i^n x_i^2$) & $Guess_{f}$ ($f(x) = \sum_i^n \frac{1}{100} x_i^2 - \cos(x)$) \\
\midrule
&    [5.0, -3.0, 7.0, 8.0] &    0.3 &    0.1 &   [0.020, -0.012, 0.029, 0.033] &    [6.150, -0.314, 6.165, 6.176] \\
&    [5.0, -3.0, 7.0, 8.0] &    0.3 &  1e-05 &   [0.000, -0.000, 0.000, 0.000] &    [6.160, -0.003, 6.160, 6.160] \\
&    [5.0, -3.0, 7.0, 8.0] &  0.003 &    0.1 &   [1.172, -0.703, 1.641, 1.876] &    [5.005, -2.999, 6.995, 7.993] \\
&    [5.0, -3.0, 7.0, 8.0] &  0.003 &  1e-05 &   [0.012, -0.007, 0.017, 0.019] &    [6.092, -0.932, 6.200, 6.267] \\
 &   [1.0, -1.0, -3.0, 2.0] &    0.3 &    0.1 &  [0.010, -0.010, -0.031, 0.020] &   [0.007, -0.007, -0.314, 0.024] \\
 &   [1.0, -1.0, -3.0, 2.0] &    0.3 &  1e-05 &  [0.000, -0.000, -0.000, 0.000] &   [0.000, -0.000, -0.003, 0.000] \\
 &   [1.0, -1.0, -3.0, 2.0] &  0.003 &    0.1 &  [0.736, -0.736, -2.207, 1.471] &   [0.995, -0.995, -2.999, 1.994] \\
 &   [1.0, -1.0, -3.0, 2.0] &  0.003 &  1e-05 &  [0.007, -0.007, -0.022, 0.015] &   [0.051, -0.051, -0.932, 0.143] \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\captionof{table}{Performance in gradient calculation for analytical and central difference approaches} 
\begin{tabular}{llllll}
\toprule
{} &   $x_{i}$  &  function &   $\nabla_{analytical}$  & $\nabla_{approximate}$ \\
\midrule
&    [5.0, -3.0, 7.0, 8.0] & $f(x) = \sum_i^n x_i^2$ &  [ 10.  -6.  14.  16.] & [ 10.  -6.  14.  16.]\\
&    [5.0, -3.0, 7.0, 8.0] & $f(x) = \sum_i^n \frac{1}{100} x_i^2 - \cos(x)$ & [-0.8589 -0.2011  0.797   1.1494] &[-0.8589 -0.2011  0.797   1.1494] \\
&    [ 1. -1. -3.  2.] & $f(x) = \sum_i^n x_i^2$ &  [ 10.  -6.  14.  16.] & [ 10.  -6.  14.  16.]\\
&    [ 1. -1. -3.  2.] & $f(x) = \sum_i^n \frac{1}{100} x_i^2 - \cos(x)$ & [ 0.8615 -0.8615 -0.2011  0.9493] & [ 0.8615 -0.8615 -0.2011  0.9493] \\
\bottomrule
\end{tabular}
\end{table}

@@

\subsubsection*{Linear Basis Function Regression}

We now consider the linear combination of basis function class regression models. We first derive the maximum-likelihood estimator for the weight vector $\theta$. Note that if we assume Gaussian errors, that is $Y \sim N(X\theta, \sigma^2)$, then our maximum-likelihood estimator is equivalent to the OLS estimator. Hence:
\begin{equation*}
	\hat{\theta}_{ML}=\hat{\theta}_{OLS}=(\Phi^T \Phi)^{-1}\Phi^T Y
\end{equation*}
where $\Phi$ denotes the polynomial basis expansion of $X$.

@@

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Meq0.png}
		\caption*{$M=0$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Meq1.png}
		\caption*{$M=1$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Meq3.png}
		\caption*{$M=3$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Meq9.png}
		\caption*{$M=9$}
	\end{minipage}
	\caption{Plots of polynomials having various orders $M$, in red, fitted to the the observed points. The green curve shows the DGF $\sin(2\pi x)$.}
\end{figure}

@@

The sum of squared errors of a linear polynomial basis model is defined by: 
\begin{equation*}
	SSE=(\Phi \theta-Y)^T(\Phi \theta-Y),         	\nabla SSE = 2\Phi^T(\Phi \theta-Y)
\end{equation*}
We tested the accuracy of this closed form solution using central differences. The analytical and numerical gradients were essentially identical across many points, even at order $M=9$. This is exactly what we want!

We then applied gradient descent to the SSE function using the analytical gradient to see if we could replicate the plots in Bishop. Optimizing our hyperparameters got increasingly difficult as we increased the order $M$. For $M=0$, we merely needed to tune the step size and we could achieve consistent convergence to the Bishop plot across many different starting values. At $M=1$, we needed to change both the step size and increase the number of maximum iterations before we consistently converged to the optimum. At $M=3$, we also had to tune the convergence threshold to be several order of magnitudes smaller than it was before, and further increase the number of max interations to get consistent convergence. For $M=9$, regardless of how we tried to tune the hyperparameters, we couldn't manage to find a set of hyperparameters that consistently converged, in fact, even scipy's BFGS optimizer failed to consistently produce a plot similar to Bishop's.

@@

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Meq0T.png}
		\caption*{$M=0$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Meq1T.png}
		\caption*{$M=1$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Meq3T.png}
		\caption*{$M=3$}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Meq9T.png}
		\caption*{$M=9$}
	\end{minipage}
	\caption{Plots of trigonometric basis expansions with various orders $M$, in red, fitted to the the observed points. The green curve shows the DGF $\sin(2\pi x)$.}
\end{figure}

@@

\subsubsection*{Implementing Ridge Regression}

Let's now see how regression behaves when we impose a regularization term. The regularization term, $\lambda$, serves to make sure we don't overfit to our training data. We implemented the closed form solution of Ridge Regression in Python, making use of Scipy and Numpy. Note that if you want to test cases where the number of features is greater than the number of data points (something that a well-trained teacher of machine should be skeptical to do), it is useful to use the pseudo-inverse, rather than the inverse to fit your parameters. We learned this the long way after debugging code for hours.

@@

Given the data in Bishop 1.4 (which can be seen below), we attempt to fit polynomials of various orders, $M$.  The plot below shows the mean squared error (MSE) that our ridge regression fit gives for values of $M$ between 1 and 5, for values of $\lambda$ ranging from $(10^0 -1)$ to $10^2$. Note that in the case where $\lambda = 0$, we have $OLS$ regression, and as $M$ increases, the model is free to overfit our data. While the MSE on training data is low, if we evaluated this on a new data point, our model would likely perform very poorly. As $\lambda$ gets higher, we penalize the model for having large $\hat{\theta}$ components. Our $MSE$s for these models are relatively large. However, these "large" MSE models are likely to perform better on a new data point!

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.48\linewidth}
                \includegraphics[width=1\linewidth]{Bishop_14_Data.png}
		\caption*{Original Bishop Data}
	\end{minipage}
	\begin{minipage}[b]{.48\linewidth}
                \includegraphics[width=1\linewidth]{MSE_Lambda_Bishop.png}
		\caption*{$MSE$($M$,$\lambda$)}
	\end{minipage}
\end{figure}

In order to avoid overfitting, while making sure that we have chosen values of $\lambda$ and $M$ that make sense, it is common to split the data into a training set and a test set. We will use the closed-form solution for ridge regression to calculate $\hat{\theta}$ on our training set. We will then use that $\hat{\theta}$ to attempt to predict values of $\mathbf{y}$ for the test data. By choosing the $\hat{\theta}$ that is trained on the training data and minimizes MSE on the test data, we are ensuring that we have not overfit our model to the data it is trained on. As one final step in the model selection process, we can check the performance of our final $\hat{\theta}$ on a validation set, to make sure that performance is still good (and we have not overfit to the data in on our test set).

@@

Let's now use this methodology on a real dataset, as opposed to a toy dataset. The dataset that we will use was compiled by Kirsztian Buza at Budapest University of Technology and Economics. It includes 280 numerical values describing a number of blog posts, such as day of week, number of parent pages, and number of comments received within 24 hours. We are hoping to predict the number of comments it will receive in the next 24 hours. This data is split into training, validation, and test datasets, so we will once again perform a grid search to find the value of $\lambda$ that minimizes $MSE$ on our validation set. In this particular case, we are not regressing a polynomial of arbitrary order, so we do not need to test different values of $M$.

We find that for very large $\lambda$ (as large as $10^6$, $MSE$ continues to decrease, although it is always close to 900. A predictor that uniformly predicts the average value of $y$, $\hat{y}$, has an $MSE$ of 1453, so the performance of our model is still better. However, we see that for values of $\lambda$ higher than $10^6$ or so, the value of the MSE increases, and eventually approaches (and eclipses) the MSE of our uniform estimator.

This isn't surprising - the blog feedback dataset is large, with 10,480 observations in the test set. If we interpret the value of $\lambda$ as the strength of our Bayesian prior on our model parameters, it makes sense that our choice of $\lambda$ does not have a substantial effect unless it is significantly larger than the size of our data. Furthermore, the inability of our model to get very low MSE for low $\lambda$ speaks both to the fact that the data has a much higher dimension than our training data, and also suggests that the simple linear model we are using here may not be a very good fit for the data.

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.48\linewidth}
                \includegraphics[width=1\linewidth]{MSE_comparison.png}
                \caption*{$MSE$ across different subsets of toy data}                
	\end{minipage}
	\begin{minipage}[b]{.48\linewidth}
                \includegraphics[width=1\linewidth]{BlogFeedbackRegressionMSE.png}
                \caption*{$MSE$($\lambda$) for blog data}
	\end{minipage}
\end{figure}

\subsubsection*{Sparse Regression with LASSO}

Let's now consider the opposite of a 10,480 point dataset - the case where we regress with only 5  training points. For this case, we might consider using LASSO, which penalizes our model with the magnitude of $\lambda$ (as opposed to its squared magnitude). Let's compare LASSO and ridge regression. We first start by determining the optimal penalty $\lambda$ that minimizes the ridge regression $MSE$ on the test set. Our optimal $\lambda$, 0.1, results in an MSE of $0.4395$. When we disable the $L_2$ regularizer, our ridge regression produces the same estimated parameters as OLS, which results in a MSE of $4.0979$. Separately, we determine that $\lambda = .9$ is the optimal LASSO penalty. This results in a $MSE$ of $0.4524$. We plot our 3 obtained fits - OLS, Ridge with $\lambda=0.1$, and LASSO with $\lambda=0.9$ - against the the true DGF.

While its seems the ridge regression fits the true DGF slightly better than LASSO, both these methods vastly outperform standard OLS. We also plot the estimated optimal weight vectors from our 3 models against the true $w$:

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.48\linewidth}
                \includegraphics[width=1\linewidth]{plot4}
	\end{minipage}
	\begin{minipage}[b]{.48\linewidth}
                \includegraphics[width=1\linewidth]{Weight_Distributions.png}
	\end{minipage}
\end{figure}

We see that the $L_1$ and $L_2$ regularizers are working as intended! Our ridge regression estimates are considerably smaller than our OLS estimates. Furthermore, our we see sparse estimation from LASSO. This sparsity results from the differences between the $L_1$ and $L_2$ penalty. In ridge regression, we want to minimize the squared error, subject to the constraint that the sum of the weights squared is less than some constant. Geometrically, we can think of this constraint as a sphere of some distance around the origin. In LASSO, we are minimizing the squared error subject to the constraint the the sum of the absolute values of the weights is less than some constant. The $L_1$ constraint can be thought of as a diamond (or regular octahedral) around the origin.

Now consider sets of weights that lead to the the same squared error. These sets can represented as ellipses with the OLS estimates at their center, as that's the point the minimizes squared error. In ridge regression, we're looking for the point where one of these ellipses touches the sphere around the origin, while in LASSO we're looking for the ellipse that touches the diamond. Since the diamond has sharper points, its fairly likely that the ellipse intersects the diamond at one of its points, which represents some weights = 0. It seems that if you expect your weights are sparse, LASSO is the correct regression method to use!
